---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# esemComp

<!-- badges: start -->
<!-- badges: end -->

esemComp helps composing ESEM-within-CFA syntaxes to be run in lavaan. It has helper functions to create target rotation matrices and to run (ESEM-like) exploratory factor analyses.

## Installation

You can install esemComp from [GitHub](https://github.com/MateusPsi) with:

``` r
# install.packages("devtools")
devtools::install_github("MateusPsi/esemComp", build_vignettes = TRUE)
```

## Example

The basic usage of the package follows the steps: make target rotation matrix (optional) > do exploratory factor analysis > compose syntax > run model in lavaan > export results (optional).

```{r example}
library(esemComp)
# use Holzinger and Swineford (1939) dataset in lavaan package
hw_data <- lavaan::HolzingerSwineford1939
hw_data <- hw_data[,c(7:15)]

#make exploratory analysis with geomin rotation
geomin_efa <- esem_efa(hw_data,3)
referents_list <- list(textual = "x5", visual = "x3", speed = "x7")
model_syntax <- syntax_composer(geomin_efa, referents_list)
writeLines(model_syntax)

# esem-within-cfa
esem_w_cfa <- lavaan::cfa(model_syntax, data = hw_data, std.lv = TRUE)

```

```{r eval=FALSE}
# save results to a text file
export_lavaan_results(esem_w_cfa)
```

Be sure to check the _esem-within-cfa_ vignette for a more comprehensive display of the package logic and capabilities:
```{r eval=FALSE}
library(esemComp)
vignette("esem-within-cfa")
```

Bug reports, feature requests or contributions are welcome! Get in touch on github or by email.
